---
title: "Assignment 2: Coding Basics"
author: "Anne Harshbarger"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1.
seq100_by4 <- seq(1,100,4); seq100_by4 #create sequence from 1-100 by 4 called "seq100_by4," 
#then call sequence to view it

#2. 
mean(seq100_by4) #calculate mean of seq100_by4
median(seq100_by4) #calculate median of seq100_by4

#3. 
mean(seq100_by4) > median(seq100_by4) #evaluate statement that the mean of the sequence is 
#greater than the median and return TRUE or FALSE

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#5 and 6.
names_v <- as.character(c("Harper", "Olivia", "Lance", "Sarah")) #Create student name vector 
#of character data
scores_v <- as.numeric(c(49, 100, 80, 77)) #Create test score vector of numeric data
pass_v <- as.logical( c("FALSE", "TRUE", "TRUE", "TRUE")) #Create pass/fail vector of logical data

#7.
student_test_results_df <- as.data.frame(cbind(names_v,scores_v,pass_v)) #Bind vectors into dataframe

#8.
names(student_test_results_df) <- c("student_names", "test_scores", "pass_fail") #Rename columns
names(student_test_results_df) #View column names to verify renaming worked

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: The data frame is able to contain different types of data. I created this data frame by binding three vectors of equal length, each containing a different type of data: character, numeric, and logical. If I had created a matrix from three vectors, all three would have needed to contain the same type of data (e.g. all numeric or all character).


10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.


11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
#10.
passing_grade <- function(score){
  
  x=score
  
  # if(x>=50) {
  #   print=TRUE
  # }
  # else {
  #   print=FALSE
  # }
  
  ifelse(x>=50, TRUE, FALSE)
  
}

#11.
results <- passing_grade(score=scores_v); results #create vector containing outputs of 
#"passing_grade" function, where input is vector "scores_v"

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: "ifelse" worked, but "if" and "else" did not; "if" and "else" (code visible as comments) was only able to run one element at a time, so the function written with this combination only determined whether the first score in the vector was a passing grade. However, "ifelse" was able to run each element on the list and return the results as another vector with length 4.


**Duke Community Standard affirmation:** I have adhered to the Duke Community Standard in completing this assignment. - Anne Harshbarger
