---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Anne Harshbarger"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A06_GLMs.Rmd") prior to submission.

The completed exercise is due on Tuesday, March 2 at 1:00 pm.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1
getwd()
setwd("C:/Users/Anne/Documents/ENV872/Environmental_Data_Analytics_2021/Assignments")
library(tidyverse)
library(agricolae)
library(lubridate)
library(ggplot2)

Lake_chem_phys <- read.csv("../Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
Lake_chem_phys$sampledate <- as.Date(Lake_chem_phys$sampledate,
                                     format = "%m/%d/%y")

#2
mytheme <- theme_bw() + theme(
  text = element_text(color = "black", size = 12),
  axis.text = element_text(color = "black", size = 12),
  legend.position = "top",
  plot.background = element_rect("white", linetype = NULL),
  panel.background = element_rect("#e9fbff", linetype = NULL), 
  axis.line = element_line(color = "gray", size = 1.25))
theme_set(mytheme) #set as default theme

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:

> Answer:
H0: Mean July temperature has no correlation with depth.
Ha: Mean July temperature is significantly correlated with depth.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
Lake_chem_phys_tidy <-
  Lake_chem_phys %>%
  mutate(Month = month(sampledate)) %>% #create month column
  filter(Month == "7") %>% #filter for July data
  select(lakename, year4, daynum, depth, temperature_C) %>% #select columns
  drop_na()

#5
temp_depth <- ggplot(data=Lake_chem_phys_tidy, 
                     aes(x = temperature_C, y = depth)) +
  geom_point(size=0.5) +
  xlim(0,35) +
  scale_y_reverse(limits = c(17.5, -0.5)) + #put depth 0 at top for readability
  geom_abline(intercept = 0, slope = 0) + #create line to show lake surface
  annotate(geom="text", x=4, y=-0.5, label = paste("Surface (Depth = 0 m)")) + 
  geom_smooth(method = lm) +
  labs(x = "Temperature (degrees C)", y = "Depth (m)", 
       title = "July temperature profile")
print(temp_depth)

```
Figure 1: Temperature profile with increasing depth for all lakes in July.

6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: Generally, temperature decreases as depth increases. The relationship between temperature and depth appears to be negative and relatively linear between the surface (0 m) and approximately 7.5 m deep, then temperature remains relatively constant between 7.5 m and the mximum depth. However, there are also fewer data points recorded for each depth after approximately 7.5 m, so this change in trend likely will not have a large impact on the linear regression.


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
temp_depth_lm <- lm(Lake_chem_phys_tidy$temperature_C ~ Lake_chem_phys_tidy$depth)
summary(temp_depth_lm)

cor.test(Lake_chem_phys_tidy$temperature_C, Lake_chem_phys_tidy$depth) 
#run cor test to confirm results

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: There is a statistically significant correlation between temperature and depth (p = 2.2 x 10^-16).  For each increase in depth of 1 meter, the temperature can be expected to drop by 1.94 degrees C. 73.9% of the variation in temperature is explained by depth. This finding is based on 1 and 9726 degrees of freedom. This finding is confirmed with a Pearson's correlation test, which also finds a statisticallly significant negative correlation.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
tempAIC <- lm(data = Lake_chem_phys_tidy, 
                temperature_C ~ depth + year4 + daynum) #set up model for AIC selection
step(tempAIC) #run AIC stepwise selection

#10
tempmodel <- lm(data = Lake_chem_phys_tidy,
                temperature_C ~ depth + year4 + daynum) #create model of best fit
summary(tempmodel) #view summary of model

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The stepwise AIC selection suggests that all three candidates for explanatory variables (depth, year4, and daynum) should be kept in the model that best fits the temperature data. The full model has the lowest AIC score at 26066; the next best model would be the one that drops year4, with an AIC score of 26070. This model explains 74.1% of the variation in temperature, which is only a slight improvement over the depth-only model, which accounted for 73.9% of the variation in temperature.



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
temp.anova1 <- aov(data = Lake_chem_phys_tidy, temperature_C ~ lakename)
summary(temp.anova1) #ANOVA as aov

temp.anova2 <- lm(data = Lake_chem_phys_tidy, temperature_C ~ lakename)
summary(temp.anova2) #ANOVA as linear model

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, based on the results of the ANOVA, at least one lake has a mean temperature that is significantly different from the other lakes (F(8,9719) = 50, p < 2x10^-16). Running the ANOVA as a linear model confirms this result - the F value, degrees of freedom, and p value are the same.



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
tempbylake <- ggplot(data=Lake_chem_phys_tidy, 
                     aes(x = depth, y = temperature_C, 
                         color = lakename)) +
  geom_point(alpha = 0.5) + #change point transparency
  ylim(0,35) + #limit temperature range
  geom_smooth(method = lm, se = FALSE) + #add line of best fit
  labs(x = "Depth (m)", y = "Temperature (C)", 
       title = "July temperatures by depth for all lakes")
print(tempbylake)

```
Figure 2: Changes in temperature with depth for all lakes in July. Color indicates lake name.

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
TukeyHSD(temp.anova1)

lake.temp.groups <- HSD.test(temp.anova1, "lakename", group = TRUE)
lake.temp.groups

temp_means <- ggplot(data = Lake_chem_phys_tidy, #plot temp distributions by lake
                     aes(x = lakename, y = temperature_C)) +
  geom_boxplot() +
  stat_summary(geom = "text", fun = max, vjust = -1, size = 3.5,
               label = c("a", "ab", "e", "de",
                         "c", "c", "de", "bc", "d")) + #label groups
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1)) + #adjust axis text for readability
  labs(x = "Lake Name", y = "Temperature (degrees C)",
       title = "July temperatures by lake")
print(temp_means)

```
Figure 3: Group distributions of temperatures for all lakes in July. Labels indicate groups as determined by HSD.test function.

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Peter Lake has the same mean temperature as Paul Lake and Ward Lake. Peter Lake has a different mean temperature from Central Long Lake, Crampton Lake, East Long Lake, West Long Lake, Hummingbird Lake, Tuesday Lake. There are no lakes with a mean temperature that is statistically distinct from all other lakes. If a lake were completely distinct, after running the HSD.test function it would be assigned to a group that was not repeated for any other lake; however, for this dataset each group occurs at least twice.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: We could also use a T-test to test the difference in mean July temperature between Peter Lake and Paul Lake, since we are only interested in comparing these two averages.

**Duke Community Standard affirmation:** I have adhered to the Duke Community Standard in completing this assignment. -Anne Harshbarger
